Safe Transferable Regions

### 2016-10-07 paper/type-inference.tex
type inference updates

\newcommand{\soln}{\eta}
\newcommand{\saturate}[1]{{#1}^*}
\newcommand{\groundc}[1]{\saturate{#1}_g}
\newcommand{\satC}{\saturate{C}}

@@

\paragraph{Constraint Solution.}
We define an \emph{assignment} $\soln$ to be a pair of functions $(\regionSubstFn,\predSubstFn)$,
where $\regionSubstFn$ is a map from $\regionVars$ to $\regionConstants$
and $\predSubstFn$ is a map from $\predVars$ to a region-constraint formula.
Such an assignment is said to be a solution to a set of constraints $C$ if
every sequent in $C$ is valid after the substitutions $\predSubstFn$ and $ \regionSubstFn$.

@@

\begin{theorem}
\label{thm:constraint-generation-sc}
Let $C$ be the set of constraints generated from a program $q$.
Let $\soln$ be any substitution for the region and predicate variables in $q$.
$\soln$ is a solution to $C$ iff $q[\soln]$ is well-typed.
\end{theorem}
\begin{proof}
The correspondence between the static semantics rules and the constraint generation
rules induces a correspondence between derivation trees produced by the
two systems.
% We do not present a formal proof of the theorem.
We can establish the following lemma inductively:
given a derivation tree $\zeta_1$ for
$\exprok {\stdcontext}{e}{\tau}{C}$
and any substitution $\soln$,
we can construct a corresponding \emph{candidate} derivation
tree $\zeta_2$ of $\hastyp {\A[\soln],\env[\soln]}{e[\soln]}{\tau[\soln]}$.
We can show that $\zeta_2$ is a valid derivation tree iff $\soln$ satisfies $C$.
(We make use of analogous lemmas for type well-formedness rules as well.)
\end{proof}

@@

\paragraph{Constraint Solver.}
Our algorithm for solving a set of constraints $C$ consists of the following steps:
\begin{enumerate}
\item We use an iterative fixed point (saturation) algorithm to compute
$\saturate{C}$, the set of all constraints logically implied by $C$.
\item Let $\groundc{C}$ denote the subset of all ground constraints
(\ie, constraints without any region variable or predicate variable)
in $\saturate{C}$. We check if $\groundc{C}$ is valid in the theory of
partial-order. If $\groundc{C}$ is valid, the solution $\soln$ is determined
as below. Otherwise, the constraint solver fails, as no valid solution exists.
\begin{itemize}
\item For any predicate variable $\varphi$, $\predSubstFn(\varphi)$ is defined
to be $\conj \{ \pi_1 \outlives \pi_2 \;|\; \pi_1, \pi_2 \in \regionConstants, \isvalid{\varphi}{\pi_1 \outlives \pi_2} \in \saturate{C} \}$.
\item For any region variable $\rho$,  the set
 $\{ \pi \in \regionConstants \:|\; \isvalid{\ell}{\rho = \pi} \in \saturate{C} \}$ 
is guarantted to contain a single element $\pi$.
We define $\regionSubstFn(\rho)$ to be $\pi$.
%\item For any region variable $\rho$,  we are guaranteed that there exists a unique
% $\pi \in \regionConstants$  such that $\saturate{C}$ contains some constraint
%with a consequent $\rho = \pi$. We define $\regionSubstFn(\rho)$ to be $\pi$.
\end{itemize}
\end{enumerate}

We compute $\saturate{C}$ by iteratively applying the following rules until a fixed
point is reached:

@@

\item (Transitivity)
$\isvalid{\ell}{\rho_1 \outlives \rho_2} \in \saturate{C}$,
$\isvalid{\ell}{\rho_2 \outlives \rho_3} \in \saturate{C}$,
$\Rightarrow$
$\isvalid{\ell}{\rho_1 \outlives \rho_3} \in \saturate{C}$

\item (Substitution)
$\isvalid{\ell}{F(\varphi)} \in \saturate{C}$,
$\isvalid{\varphi}{\phi} \in \saturate{C}$
$\Rightarrow$ $\isvalid{\ell}{F(\phi)} \in \saturate{C}$

@@

Point~\ref{item:context} makes use of the special structure of the generated
constraints: namely that $\phictxt$  is a conjunction of outlives-constraints of the
form $\pi \outlives \pi_s$ where $\pi_s$ is a static region identifier and $\pi$ is a
region constant. As a result, we can show that $\isvalid{\varphi \conj \phictxt}{\pi_i \outlives \pi_j}$
iff $\isvalid{\varphi}{\pi_i \outlives \pi_j}$ or $\isvalid{\phictxt}{\pi_i \outlives \pi_j}$.

\paragraph{Properties of generated constraints}
Our algorithm makes use of some special properties of $C$ guaranteed by
our constraint-generation algorithm, which we now describe.

For any predicate variable $\varphi$ used in $C$, $C$ has exactly one constraint
of the form $\tywf{\Delta}{\varphi}$. We will refer to this $\Delta$ as
$\predDeltaMap{\varphi}$.

\begin{lemma}
Let $C$ be the set of constraints generated by our algorithm.
Every abduction constraint in $C$ is decomposable.
\end{lemma}

\begin{lemma}
  Consider any decomposable constraint  $\isvalid{\varphi \conj \phi}{\pi_i \outlives \pi_j}$
where both $\pi_i$ and $\pi_j$ are region constants.
(a) If $\{ \pi_i, \pi_j \} \subseteq \predDeltaMap(\varphi)$,
then $\soln$ satisfies $C \cup \{ \isvalid{\varphi \conj \phi}{\pi_i \outlives \pi_j} \}$
iff
$\soln$ satisfies $C \cup \{ \isvalid{\varphi}{\pi_i \outlives \pi_j} \}$
  (b) If $\{ \pi_i, \pi_j \} \not \subseteq \predDeltaMap(\varphi)$,
$\soln$ satisfies $C \cup \{ \isvalid{\varphi \conj \phi}{\pi_i \outlives \pi_j} \}$
iff
$\soln$ satisfies $C \cup \{ \isvalid{\phi}{\pi_i \outlives \pi_j} \}$
\end{lemma}

\paragraph{Algorithmic Aspects.}
The iterative process described above is a standard fixed point computation.
We can encode this computation using a set of Datalog rules,
and compute the solution using any Datalog engine.
Alternatively, algorithms for context-free reachability can be adapted
to solve the constraints, as explained below.

Given a set $S$ of outlives constraints, we define the directed
graph $G(S)=(V(S),E(S))$ as follows.
Every distinct region identifier $\rho$ in $S$ is represented by a vertex,
which we will also refer to as $\rho$.
Every outlives constraint $\rho_1 \outlives \rho_2$ is represented by
an edge from $\rho_1$ to $\rho_2$.
Recall that an equality constraint $\rho_1 = \rho_2$ is shorthand
for the pair of constraints $\rho_1 \outlives \rho_2$ and
$\rho_2 \outlives \rho_1$: thus, an equality constraint is represented
by a pair of edges.
%
It is easy to see that $S \models \rho_1 \outlives \rho_2$ iff
there exists a path from $\rho_1$ to $\rho_2$ in $G(S)$.
%
Thus, simple graph reachability (transitive closure) algorithms can be
used to identify all the logical consequences of a set of outlives
constraints.

Given a set of context-free validity constraints, we can generalize the
above approach and reduce the problem to one of context-free reachability
in graph~\cite{Reps:Reachability} (as usual for context-sensitive interprocedural analysis).
%
Algorithms for context-free reachability can also be adapted to incorporate
the above treatment of context constraints $\phictxt$.
%
Further details omitted.

@@

Correctness follows by establishing:
\begin{theorem}
(a) For any assignment $\soln$, $\soln$ satisfies $C$ iff $\soln$ satisfies $\satC$.
(b) $\satC$ is satisfiable iff $\groundc{C}$ is valid.
(c) For any region variable $\rho$, $\satC$ contains some constraint of the
form $\isvalid{\ell}{\rho = \pi}$.
\end{theorem}

@@

\begin{theorem}
\label{thm:constraint-solver-sc}
Let $C$ be the set of constraints generated from some $p \in \absof{\FB}$.
If the constraint solver returns Some($\soln$), then $\soln$ is a valid solution
for $C$.
Conversely, if $C$ has some valid solution, then the constraint solver will return
a value of the form Some($\soln$).
\end{theorem}

@@

\subsection{Discussion}

We now state the soundness theorem:

@@

\begin{enumerate}
\item Our technique for determining the set of region parameters for a recursive class
incorporates a specific heuristic, namely that the recursive occurrence of the class has
the same parameters, in the same order, as the class itself. This technique will fail, for
example. if the program uses a recursive list type whose elements alternatively come
from two different regions. Such a program would require the following elaboration,
which is beyond the scope of our approach:
% , $R_1$ and $R_2$.
% both of which are region parameters of the class.
\begin{tabular}{ccc}
\begin{minipage}{0.3\linewidth}
\begin{codejava}
class List
  $\extends$ $\ObjZ$
{
  $\ObjZ$ data;
  List next;
}
\end{codejava}
\end{minipage}
&
$\Rightarrow$
&
\begin{minipage}{0.5\linewidth}
\begin{codejava}
class List $\langle \rho_0, \rho_1, \rho_2 \; | \; \varphi \rangle$
  $\extends$ $\ObjZ \langle \rho_0 \rangle$
{
  $\ObjZ \langle \rho_1 \rangle$ data;
  $\C{List} \langle \rho_0, \rho_2, \rho_1 \rangle$ next;
}
\end{codejava}
\end{minipage}
\end{tabular}

@@

\begin{itemize}
  \item for function-calls: need basic typing from core language to introduce region parameters
  \item discuss computation of least-solution to constraints?
  \item use of a set to represent conjunctions
  \item equality is shorthand for two inequalities
  \item use of alpha-renaming, and its use to discharge not-in contraints;
also, incorporate this into first stage rather than second (in Fig. 7).
  \item Note that we can also clarify that the well-formedness constraints are
    syntactic sugar from basic set constraints
  \item Prove: Progress: Every region-variable will be unified with some constant.
  \item Notation: should we use a different notation to distinguish a validity constraint
    from actual entailment?
\end{itemize}

>>>>

\newcommand{\soln}{\eta}
\newcommand{\saturate}[1]{{#1}^*}
\newcommand{\groundc}[1]{\saturate{#1}_g}
\newcommand{\consOf}[1]{\textsc{GenConstraint}(#1)}
\newcommand{\satC}{\saturate{C}}

@@

\paragraph{Constraint Solution.}
We define an \emph{assignment} $\soln$ to be a pair of functions $(\regionSubstFn,\predSubstFn)$,
where $\regionSubstFn$ is a map from $\regionVars$ to $\regionConstants$
and $\predSubstFn$ is a map from $\predVars$ to a region-constraint formula.
Such an assignment is said to \emph{satisfy} a set of constraints $C$ if
every sequent in $C$ is valid after the substitutions $\predSubstFn$ and $ \regionSubstFn$.
We say that $C$ is \emph{satisfiable} if it has a satisfying assignment.

@@

\begin{theorem}
\label{thm:constraint-generation-sc}
Let $C = \consOf{q}$.
Any assignment $\soln$ (for the region and predicate variables in $q$)
satisfies $C$ iff $q[\soln]$ is well-typed.
\end{theorem}
\begin{proof}
The correspondence between the static semantics rules and the constraint generation
rules induces a correspondence between derivation trees produced by the
two systems.
% We do not present a formal proof of the theorem.
We can establish the following lemma inductively:
given a derivation tree $\zeta_1$ for
$\exprok {\stdcontext}{e}{\tau}{C}$
and any substitution $\soln$,
we can construct a corresponding \emph{candidate} derivation
tree $\zeta_2$ of $\hastyp {\A[\soln],\env[\soln]}{e[\soln]}{\tau[\soln]}$.
We can show that $\zeta_2$ is a valid derivation tree iff $\soln$ satisfies $C$.
(We make use of analogous lemmas for type well-formedness rules as well.)
\end{proof}

@@

% \paragraph{Properties of generated constraints}
Our constraint solver makes use of some special properties of the set of constraints $C$
generated by our algorithm, which we now describe.

For any predicate variable $\varphi$ used in $C$, $C$ has exactly one constraint
of the form $\tywf{\Delta}{\varphi}$. We will refer to this $\Delta$ as
$\predDeltaMap(\varphi)$.

The formalism below is motivated by the following intuition.
Suppose some construct in a method's body requires an outlives-constraint
$\pi_1 \outlives \pi_2$ to hold. If both $\pi_1$ and $\pi_2$ are region parameters
of the method, then we can incorporate the requirement $\pi_1 \outlives \pi_2$ 
as part of the method's precondition ($\varphi$). Otherwise, the outlives-constraint
$\pi_1 \outlives \pi_2$ must be established by the context within the method
(\eg, by the \code{letregion} constructs in the method).
We now generalize this idea to the setting of our abstract constraint-satisfication
problem.

\begin{lemma}
  \label{lemma:gc-is-decomposable}
  Let $C = \consOf{q}$. Every abduction constraint in $C$ is decomposable.
\end{lemma}

\begin{lemma}
  \label{lemma:decomposition}
  Consider any decomposable constraint  $\isvalid{\varphi \conj \phi}{\pi_i \outlives \pi_j}$
where both $\pi_i$ and $\pi_j$ are region constants.
(a) If $\{ \pi_i, \pi_j \} \subseteq \predDeltaMap(\varphi)$,
then $\soln$ satisfies $C \cup \{ \isvalid{\varphi \conj \phi}{\pi_i \outlives \pi_j} \}$
iff
$\soln$ satisfies $C \cup \{ \isvalid{\varphi}{\pi_i \outlives \pi_j} \}$
  (b) If $\{ \pi_i, \pi_j \} \not \subseteq \predDeltaMap(\varphi)$,
$\soln$ satisfies $C \cup \{ \isvalid{\varphi \conj \phi}{\pi_i \outlives \pi_j} \}$
iff
$\soln$ satisfies $C \cup \{ \isvalid{\phi}{\pi_i \outlives \pi_j} \}$
\end{lemma}

\begin{proof}
  Straightforward.
\end{proof}

\paragraph{Constraint Solver.}
Our algorithm for solving a set of constraints $C$ consists of the following steps:
\begin{enumerate}
\item We use an iterative fixed point (saturation) algorithm to compute
$\saturate{C}$, the set of all constraints logically implied by $C$.
\item Let $\groundc{C}$ denote the subset of all ground constraints
(\ie, constraints without any region variable or predicate variable)
in $\saturate{C}$. We check if $\groundc{C}$ is valid in the theory of
partial-order. If $\groundc{C}$ is valid, the solution $\soln$ is determined
as below. Otherwise, the constraint solver fails, as no valid solution exists.
\begin{itemize}
\item For any predicate variable $\varphi$, $\predSubstFn(\varphi)$ is defined
to be $\conj \{ \pi_1 \outlives \pi_2 \;|\; \pi_1, \pi_2 \in \regionConstants, \isvalid{\varphi}{\pi_1 \outlives \pi_2} \in \saturate{C} \}$.
\item For any region variable $\rho$, we define $\regionSubstFn(\rho)$ to be any element of the
set $\{ \pi \in \regionDeltaMap(\rho) \:|\; \isvalid{\ell}{\rho = \pi} \in \saturate{C} \}$ 
if this set is non-empty.
Otherwise, the solver fails.
%\item For any region variable $\rho$,  we are guaranteed that there exists a unique
% $\pi \in \regionConstants$  such that $\saturate{C}$ contains some constraint
%with a consequent $\rho = \pi$. We define $\regionSubstFn(\rho)$ to be $\pi$.
\end{itemize}
\end{enumerate}

\paragraph{Step 1.}
We compute $\saturate{C}$ by iteratively applying the following rules until a fixed
point is reached:

@@

\item (Transitivity)
  \label{item:transitivity}
$\isvalid{\ell}{\rho_1 \outlives \rho_2} \in \saturate{C}$,
$\isvalid{\ell}{\rho_2 \outlives \rho_3} \in \saturate{C}$,
$\Rightarrow$
$\isvalid{\ell}{\rho_1 \outlives \rho_3} \in \saturate{C}$

\item (Substitution)
  \label{item:subst}
$\isvalid{\ell}{F(\varphi)} \in \saturate{C}$,
$\isvalid{\varphi}{\phi} \in \saturate{C}$
$\Rightarrow$ $\isvalid{\ell}{F(\phi)} \in \saturate{C}$

@@

Point~\ref{item:context} makes use of Lemmas~\ref{lemma:gc-is-decomposable} and~\ref{lemma:decomposition}.
% the special structure of the generated
% constraints: namely that $\phictxt$  is a conjunction of outlives-constraints of the
% form $\pi \outlives \pi_s$ where $\pi_s$ is a static region identifier and $\pi$ is a
% region constant. As a result, we can show that $\isvalid{\varphi \conj \phictxt}{\pi_i \outlives \pi_j}$
% iff $\isvalid{\varphi}{\pi_i \outlives \pi_j}$ or $\isvalid{\phictxt}{\pi_i \outlives \pi_j}$.

\paragraph{Step 2.}
Checking the validity of a ground constraint is straightforward.
A ground constraint is of the form $\isvalid{\conj_{i \in I} \ell_i}{r}$, where each
$\ell_i$ and $r$ is an outlives-constraint.
Let $D$ denote $\{ \isvalid{}{\ell_i} \;|\; i \in I \}$.
The given constraint is valid iff $r$ belongs to $\saturate{D}$.

\paragraph{Algorithmic Aspects.}
Note that checking the validity of a ground constraint can be realized using
a simple graph reachability algorithm.
%
Given a set $S$ of outlives constraints, define the directed
graph $G(S)=(V(S),E(S))$ as follows.
Every distinct region identifier $\rho$ in $S$ is represented by a vertex,
which we will also refer to as $\rho$.
Every outlives constraint $\rho_1 \outlives \rho_2$ is represented by
an edge from $\rho_1$ to $\rho_2$.
(Recall that an equality constraint $\rho_1 = \rho_2$ is shorthand
for the pair of constraints $\rho_1 \outlives \rho_2$ and
$\rho_2 \outlives \rho_1$: thus, an equality constraint is represented
by a pair of edges.)
%
It is easy to see that $\isvalid{\conj S}{\rho_1 \outlives \rho_2}$ iff
there exists a path from $\rho_1$ to $\rho_2$ in $G(S)$.
%
Thus, a simple graph reachability algorithm can be used to check the validity of
ground constraints.
% all the logical consequences of a set of outlives constraints.

This idea generalizes. Consider a set $C$ of constraints that contains no
constraint with an antecedent $\varphi \conj \phi$ where $\phi$ is a non-empty conjunction. 
Then, the algorithm for computing the closure $\satC$ does not need to use
rule~\ref{item:context} (for Abduction Decomposition). The closure in this
case can be computed using context-free reachability in graphs~\cite{Reps:Reachability}
(as usual for context-sensitive interprocedural analysis).

Algorithms for context-free reachability can be adapted to incorporate
the Abduction Decomposition step.
% Further details omitted.
Alternatively, the iterative process described above is a standard fixed point computation.
We can encode this computation using a set of Datalog rules,
and compute the solution using any Datalog engine.

@@

Correctness follows by establishing:
\begin{theorem}
(a) Any assignment $\soln$ satisfies $C$ iff $\soln$ satisfies $\satC$.
(b) $\satC$ is satisfiable iff $\groundc{C}$ is valid.
(c) For any region variable $\rho$, $\satC$ contains some constraint of the
form $\isvalid{\ell}{\rho = \pi}$.
\end{theorem}

@@

\begin{theorem}
\label{thm:constraint-solver-sc}
Let $C = \consOf{q}$.
(Soundness) If the constraint solver succeeds, it will return an assignment $\soln$ 
that satisfies $C$.
(Completeness)
Conversely, if $C$ is satisfiable, then the constraint solver will succeed.
\end{theorem}

@@

\subsection{Soundness and Completeness: Discussion}

We now state the soundness theorem, which follows from
Theorems~\ref{thm:constraint-generation-sc} and~\ref{thm:constraint-solver-sc}
as well as the following observation about the elaboration phase.
For any program $p \in \absof{\FB}$, the elaboration phase produces a $q$
such that for any assignment $\soln$, we have $\absof{q[\soln]} = p$.

@@

\begin{enumerate}
\item We determine the set of region parameters for a recursive class
using the heuristic that a recursive occurrence of the class has
the same parameters, in the same order, as the class itself.
This heuristic fails, for example, if the program uses a recursive list type whose
elements alternatively come from two different regions. Such a program would require
the following elaboration, which is beyond the scope of our approach:

\begin{tabular}{ccc}
\begin{minipage}{0.35\linewidth}
\begin{codejava}
class List $\extends$ $\ObjZ$
{
  $\ObjZ$ data;
  List next;
}
\end{codejava}
\end{minipage}
&
$\Rightarrow$
&
\begin{minipage}{0.65\linewidth}
\begin{codejava}
class List $\langle \rho_0, \rho_1, \rho_2 \; | \; \varphi \rangle$ $\extends$ $\ObjZ \langle \rho_0 \rangle$
{
  $\ObjZ \langle \rho_1 \rangle$ data;
  $\C{List} \langle \rho_0, \rho_2, \rho_1 \rangle$ next;
}
\end{codejava}
\end{minipage}
\end{tabular}

@@

\begin{itemize}
  \item explain: non-determinism in defining region-variable binding is immaterial.
  \item explain lemma 2's use case (a) and (b)
  \item notation: decomposable is dependent on C
  \item cleanup: syntax of constraints (region-constraint, validity constraint, etc.)
    and later usage (abduction constraint)
  \item adjust spacing in RULEs
  \item for function-calls: need basic typing from core language to introduce region parameters
  \item discuss computation of least-solution to constraints?
  \item use of a set to represent conjunctions
  \item equality is shorthand for two inequalities
  \item use of alpha-renaming, and its use to discharge not-in contraints;
also, incorporate this into first stage rather than second (in Fig. 7).
  \item Note that we can also clarify that the well-formedness constraints are
    syntactic sugar from basic set constraints
  \item Prove: Progress: Every region-variable will be unified with some constant.
  \item Notation: should we use a different notation to distinguish a validity constraint
    from actual entailment?
  \item update abstract/intro to stress aspects of constraint solver
  \item Change static semantics to use method precondition, without class precondition
\end{itemize}
